---
title: "Midewin_Diversity_Comp.Results.Figures"
author: "Marissa Zaricor"
date: "2026-05-11"
output: 
  html_document:
    smart: no
    theme: readable
    float: yes
editor_options: 
  chunk_output_type: console
mainfont: Times New Roman 
mathfont: Times New Roman
fontsize: 12pt
---
```{r setup, include=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE)
#hide code, warnings, and messages from final ouptput document
```

#  {.tabset .tabset-pills .tabset-fade}

## Set up and Overview

### Running Title: Management modifies consumer top-down control of plant diversity in grassland restoration


**Background:** Quantifying the independent and interactive effects of management (fire, mow, herbicide, none) and consumers (bison, small mammals, no mammals) on plant community composition and diversity (species richness, Shannon diversity, and evenness) during tallgrass prairie restoration  

**Author:** Marissa Zaricor

**Co-authors:** Lauren Sullivan

**Created:** 2025-05-16

**Files:**

1. *Midewin_cover_clean.csv* - contains plant cover, codes, taxon, etc.

2. *Midewin_Aug-Aug_Grazing_Influence_22_24.csv*- bison raster data for total (sum by year) and weighted total (sum of weighted by month/season) but aug year-1 to aug year current. (last updated Jan 2025)

**Last Updated:** 2026-06-05

#### Libraries, Files, Setup
```{r echo= FALSE}
# setup
rm(list=ls()) # clean up workspace 


library(tidyverse) #data cleaning and visualization
library(vegan)  # for plant comm ecology ( nmds,permanova, diversity,etc)
library(car)     # For VIF collinearity check
library(ggplot2) #visualization
library(ggvegan) # an extension of ggplot2 designed for comm analyses using vegan package
library(goeveg) #NMDS stress test
#library(lme4)    # For LMM and GLMM
library(glmmTMB) # For advanced mixed models
#library(mgcv)    # For GAMs
library(emmeans) # For post-hoc analysis
#library(glmm) # glmms 
library(MuMIn) #for model comparison and AIC 
library(multcomp) #post hoc tests (Tukey HSD, GLHT)
library(performance) # diagnostics, AIC 
library(DHARMa) #diagnostics checks 
#devtools::install_github("pmartinezarbizu/pairwiseAdonis/pairwiseAdonis")
library(pairwiseAdonis)# for pairwise permanova tests 
#library(BiodiversityR) # plot ordi-elippises in ggplot
library(ggpubr) #similar and compatible with ggplot for editing plots for publication
library(ggordiplots) # helps with ordination plots using vegan and integrating to ggplot
library(knitr) # help with merging data to tables 
library(kableExtra) # formatting nicer tables 
library(extrafont) #fonts for publication
#library(indicspecies) #for inidcator species analysis 

```

Code for ggplot themes for publications
```{r}
#ggplot theme for publication using ggpubr
text_theme<- function(base_size=12, font_family="Arial") {
    theme(text=element_text(family = font_family, size = base_size), 
          
        plot.title=element_text(size=base_size+1, hjust = 0.5,face = "bold"), 
        
        axis.title=element_text(size = base_size+1, color = "#0F0F0F"), #see if this is better than bold
        axis.title.x =element_text(margin=margin(t=10)),#want to move the titles slightly
        axis.title.y = element_text(margin=margin(r=10)), #same moving away from text
        
        axis.text=element_text(size = base_size-1), #will use default "#333333" light grey
        
        legend.title=element_text(size=base_size+1, color= "#0F0F0F"), #darker but not bold
        legend.text=element_text(size = base_size-1), 
        strip.text=element_text(size = base_size))
        }



#ggsave (filename= "Figure1.png" or tiff etc)

#single column: 
# ggsave ("Fig1.png", plot, width= 3.5, height=5, units = "in", dpi = 600) 

#double column: 
#  ggsave ("Fig1.png", plot, width= 7.3, height=5, units = "in", dpi = 600) 
#height here can be flexible ~ 4-6

#height can be flexible usually 4-6 inches, max is ~9



## coding figures

# to help with plotting my points that were shifting
dodge_width<- 0.6 # will try this value first for horizontal grouping
pos_dodge <- position_dodge(width = dodge_width) # will help with setting this as standard
pos_jitter <- position_jitterdodge(jitter.width = 0.1, dodge.width = dodge_width, seed = 123) # jitter will help reduce over lapping points, but we don't want them to far apart (use width to adjust the left/right shifts) and jitterdoge combines both to separate groups and spread about our points
pos_jitter2 <- position_jitterdodge(jitter.width = 0.2, dodge.width = dodge_width, seed = 123)
```



## Data clean

#### Data cleaning

<!-- ## get rid of this and just provide clean data. -->
<!-- ```{r} -->

<!-- cover<- read.csv("Midewin_Cover_21_24.csv") -->

<!--  cover_clean <- cover %>% mutate(year = as.factor(year),block= as.factor(block), plot= as.factor(plot), cover = as.numeric(cover)) %>% -->
<!--        filter(!(is.na(code) & (is.na(cover) | cover == 0))) # remove rows where code is missing or blank AND cover is na -->

<!--  cover_clean<-cover_clean[, -c(15)] #remove last column with notes -->

<!--  #double check values we don't want any negatives or over 100 (though totals can be) -->
<!--  #cover_clean %>% summarise(cover_min = min(cover, na.rm = TRUE), cover_max= max(cover, na.rm = TRUE)) #o to 100 only so thats good -->


<!--  #Change herbivory levels for XX to be our reference, control is ref by default -->

<!--  cover_clean$herbivory_trt<-factor(cover_clean$herbivory_trt, levels=c("XX", "SX", "SB")) -->

<!-- final <- cover_clean %>% -->
<!--     filter(bison_pasture=="yes", year=="2023" | year == "2024") -->

<!-- cover_clean <- final -->


<!-- #write_csv(final, "Midewin_cover_clean.csv") -->
<!-- ``` -->



```{r}

## set working directory to source file location

cover_clean<- read.csv("Midewin_cover_clean.csv")
cover_clean$herbivory_trt<-factor(cover_clean$herbivory_trt, levels=c("XX", "SX", "SB")) #relevel herbivory


graze_Aug <-read_csv("Midewin_Aug-Aug_Grazing_Influence_22-24.csv")



```












## Species Richness 


#### Calculate species richness 
```{r}
#calculate species richness 
specRich1<- cover_clean %>% 
  filter(live==1) %>%  # remove anything not a plant
  filter(unknown==0) %>%  # remove any unknowns
  filter(code !="" & !is.na(code)) %>%  #remove any empty rows in  code column
  group_by(year, block, plot, management_trt, herbivory_trt, bison_pasture) %>%  #do not include other columns here for now
  summarise(richness=n_distinct(code), .groups = "drop") #counts all distinct species per plot

```

#### Add in bison covariate data
```{r}
#plot and year need to be factors in order to merge with spec rich df
specRich1$plot<-as.factor(specRich1$plot)
specRich1$year<-as.factor(specRich1$year)
graze_Aug$plot<-as.factor(graze_Aug$plot)
graze_Aug$year<-as.factor(graze_Aug$year)

# join bison cov df and specrich df into new single df
specRich1<- specRich1 %>% left_join(graze_Aug %>% dplyr:: select(plot, year, Weighted_Grazing_Influence), by=c("plot", "year"))



#fill in missing or zero graze influence values
#each block/year has one assigned bison influence level, needs to be assigned to all plots within each block
specRich1 <-specRich1 %>% group_by(block, year) %>% 
  mutate(block_graze_value =first(Weighted_Grazing_Influence[!is.na(Weighted_Grazing_Influence)&Weighted_Grazing_Influence !=0]), Weighted_Grazing_Influence=ifelse(is.na(Weighted_Grazing_Influence)|Weighted_Grazing_Influence==0,block_graze_value, Weighted_Grazing_Influence)) %>% 
  ungroup() %>% 
  dplyr::select(-block_graze_value)

# replace any missing values with zero, these represent plots/years with no recorded bison influence
specRich1<-specRich1 %>% mutate(Weighted_Grazing_Influence=ifelse(is.na(Weighted_Grazing_Influence), 0, Weighted_Grazing_Influence))


```


#### Fit the model 
```{r,results='hide', fig.show='hide'}

#Main results 2024 so remove east and other years
rich_W24<- specRich1 %>% filter(year=="2024") 



#fitting the model, guassian 
glmm_richW24<-glmmTMB(richness~herbivory_trt*management_trt+(1|block/plot), data = rich_W24, family = gaussian)
summary(glmm_richW24)
#bison covariate is not significant and will be removed from the model

#Results: significnat interaction and main effects 

#AIC 321

#use wald chi sqaaure tests to show our approx. entire effects, using type = III because better for our interactions structure
anova_w24<-Anova(glmm_richW24, type="III")
anova_w24


```


#### Check assumptions

```{r,results='hide', fig.show='hide'}
#check assumptions
sim_glmm_W24<-simulateResiduals(glmm_richW24)
plot(sim_glmm_W24) #QQ plot is alright and residual vs predicted looks good
testResiduals(sim_glmm_W24) 
#KS test p>0.05, residuals are uniform, model is good fit
#dispersion test came out okay
```




####  r2 
```{r,results='hide', fig.show='hide'}
#what is the explained variance of our mixed effects model? 
r2(glmm_richW24) # uses performance package 

#marginal r2 (var explained by fixed effects only) = 0.583
#conditional r2 (var explained by fix and random) = .743
```

#### Post-hoc comparisons
```{r}
#herbivory within each management
emm_richW24<- emmeans(glmm_richW24, ~herbivory_trt|management_trt, df = 57) #separates for each level of management trt, matching DF to that produced in glmmTMB() summary model to not get Inf.
pairs_w24<-contrast(emm_richW24, "pairwise", adjust="tukey")
pairs_w24

 
#management within each herbivory treatment
emm_richW24m<-emmeans(glmm_richW24, ~management_trt|herbivory_trt, df = 57) #separates for each level of herbiovry  trt
pairs_w24m<-contrast(emm_richW24m, "pairwise", adjust="tukey")
pairs_w24m 

```



#### Fig. 2 -  Data visualization of Species Richness 2024 West Pasture 
```{r}
#use emmeans for estimated marginal means with 95% CI
#estimates are for each consumer treatment within each management treatment 
 emm_rich24 <-emmeans(glmm_richW24,  ~herbivory_trt|management_trt, df = 57) #  matching DF to that produced in glmmTMB() summary model to not get Inf.
 emm_df <-as.data.frame(emm_rich24)

#need a data frame with our pairwise contrasts to extract significance values  to plot
richw24_pairsdf<-as.data.frame(summary(pairs(emm_rich24, by="management_trt", adjust="tukey"))) 


#will use this data frame with ggpbur package to add to pvalues to ggplot
#keeps only significant pairwise comparisons
#contrast columns are split into group 1 and group 2 for brackets between consumer groups
pval_df<- richw24_pairsdf %>% filter(p.value <0.05) %>% # only get significant pairwise difference 
  mutate(group1 = str_split_fixed(contrast, " - ", 2)[,1], #this splits apart my contrasts SB - XX into groups
         group2= str_split_fixed(contrast, " - ",2)[,2],
         p.label=case_when( p.value<0.001 ~ "***", p.value<0.01~"**", p.value<0.05~"*"))  # assigns asterisks for pvalues 

# these are our factor levels and helps with x -axis bracket positions and keeps the order I want
herbivory_levels<-c("XX", "SX", "SB")
management_levels<- c("control", "fire", "herbicide", "mow")

#contrast groups are now made into factors for significance brackets 
#have to do a psuedo manual position becuase we have dodging, this may need to be adjusted further
pval_df<-pval_df %>% mutate(group1=factor(group1, levels = herbivory_levels), #asociates groups with herbivory levels for sb,sx,xx
        group2=factor(group2, levels = herbivory_levels), 
        management_trt=factor(management_trt, levels = management_levels), 
        xnum=as.numeric(management_trt), # calculate the x position for each treatment
        x.position1=xnum-dodge_width/2+(as.numeric(group1)-1)* (dodge_width/2), #should aligns each x position correctly
        x.position2=xnum-dodge_width/2+(as.numeric(group2)-1)* (dodge_width/2),)

#need our brackets redefined by y position for no overlap 
#need to be stacked
pval_df <- pval_df %>% group_by(management_trt) %>%  arrange(p.value) %>% #order for contrast significance
  mutate(contrast_index=row_number(), #this helps with stacking brackets 
        base_y = max (emm_df$upper.CL)+0.7, #this will adjust our height of our brackets, may need to modify this
        y.position= base_y + (contrast_index -1) *0.7) %>% #this staggers our brackets on the y
  ungroup()


#custom colors for treatments 
herbivory_colors <- c("SB" = "#440154FF", "SX" = "#35B779FF","XX" = "#31688EFF")
management_colors <- c("control" = "#000000","fire" = "#CC79A7", "herbicide" = "#E69F00","mow"= "#0072B2")


#we want to plot both raw and model estimated means 
#"transparent" = raw plot level, and larger and error bars show our emm =/- 95% CIs
ggplot()+
  # raw data points that are transparent
  geom_jitter (data=rich_W24, aes(x=management_trt, y=richness, color=herbivory_trt), size=2, alpha=0.3, shape=16, position = pos_jitter2)+ 
  # group means with CI bars from emmeans data 
  geom_point(data=emm_df, aes(x=management_trt,y=emmean, color=herbivory_trt ), position = pos_dodge, size=3)+
  #CI for error bars
  geom_errorbar(data = emm_df, aes(x=management_trt, ymin = lower.CL, ymax = upper.CL, color=herbivory_trt), width=0.2, position = pos_dodge)+
  theme_classic2()+
  text_theme(base_size = 12) + #use our text theme from above
  theme(legend.position = "bottom", plot.margin = margin(t=10, r=35, b=10, l=10))+ #need extra space for axis text
  coord_cartesian(clip = "off")+ #additional help so ggplot doesn't cut off any wording when saving 
  scale_color_manual(values = herbivory_colors,labels= c( "XX"= "No Mammals", "SX"= "Small Mammals", "SB"= "Bison + Small Mammals"))+
  labs( x= "Management Treatment", y= "Species Richness (mean +/- 95% CI)", color= "Consumer Treatment")+
  scale_x_discrete(labels = c("control"="Control", "fire"="Fire", "mow"="Mow", "herbicide"="Herbicide"))+
  stat_pvalue_manual(data = pval_df, label = "p.label", xmin = "x.position1", xmax = "x.position2", y.position = "y.position", tip.length = 0.01, bracket.size = 0.8, color = "black", size = 6)
  
#save the figure for high resolution
ggsave ("Fig2_Richness2024.png",  width= 7.3, height=5, units = "in", dpi = 600) 
```


Table: GLMM Spec Rich 2024
```{r}

#model summary
glmm_W24sum<-summary(glmm_richW24)

#make into df and round up 
glmm_W24_df <- as.data.frame(glmm_W24sum$coefficients$cond) %>% tibble::rownames_to_column("Term") %>% 
  mutate(across(c(Estimate, `Std. Error`, `z value`, `Pr(>|z|)`), ~round(.x, 3)))

#Model Summary
glmm_W24_df

```

Table 1: Anova Spec Rich 2024
```{r}

#convert to df, keep column names and round up
anova_w24_df<-as.data.frame(anova_w24) %>% rownames_to_column("Term") %>% rename(Chisq = Chisq, Df= Df, p_value= `Pr(>Chisq)`)

#Anova summary
anova_w24_df

#export to CSV 
#write.csv(anova_w24_df, "Table 1 Spec Rich Anova.csv", row.names = FALSE)
```

Table: Post Hoc Comparisons Spec Rich 2024
```{r}
#extract data for post hoc comparisons tables 
#pairs_w24 #pairs_w24m

#compares herbivory treatments within each management treatment
pairs_w24_df<-as.data.frame(pairs_w24) %>% 
  #rename(Management_Treatment = management_trt, Contrast = contrast, Estimate = estimate, SE=SE, Df= df, t_ratio = t.ratio, p_value= p.value) %>% 
  mutate(across(c(estimate, SE, t.ratio), ~round(.x, 3)), p.value=round(p.value, 4))

#write.csv(pairs_w24_df, "Table_S1_PostHoc_HerbMgmt_Rich24.csv", row.names = FALSE)

#compares management treatments to each other within herbivory treatments 
pairs_w24m_df<-as.data.frame(pairs_w24m) %>% rename( Contrast = contrast, Estimate = estimate, SE=SE, Df= df, t_ratio = t.ratio, p_value= p.value) %>% 
  mutate(across(c(Estimate, SE, t_ratio), ~round(.x, 3)), p_value=round(p_value, 4))
pairs_w24m_df


```








## Richness Delta Model (23-24)


#### Calculate the difference from 2023 to 2024 in specrich
```{r,results='hide', fig.show='hide'}
#filter data to only bison pasture and years 23-24
rich23_24<-specRich1 %>% filter(bison_pasture=="yes", year %in% c("2023", "2024"))


# rearrange and calculcate difference between years
rich_diff <- rich23_24 %>% 
  arrange(block, plot, year) %>% #makes sure 2023 is before our 2024 rows for each plot
  group_by(block, plot, management_trt, herbivory_trt) %>% #keeps our unique combos
  filter(n() == 2) %>% #only plots with data in both years are used
  mutate(diff_richness = richness - lag(richness)) %>% #this calculates our differences
  filter(!is.na(diff_richness)) %>% ungroup #remove any nas
```


#### Fit the model 
```{r,results='hide', fig.show='hide'}
diff_glmm1<- glmmTMB(diff_richness~herbivory_trt*management_trt+ (1|block/plot), data=rich_diff, family = gaussian)
summary(diff_glmm1)

anova_diff<-Anova(diff_glmm1, type="III")
#bison covariate was not significant and dropped from the model AIC 312
#only herbivory trt significant (p=0.028) 
```



#### Check assumptions

```{r,results='hide', fig.show='hide'}
#check assumptions
sim_glmm_diff1<-simulateResiduals(diff_glmm1)
plot(sim_glmm_diff1) #QQ looks alright, no problems in residual/predicted plots
testResiduals(sim_glmm_diff1) #KS good, no sign dispersion, maybe one outlier but n.s. 
check_model(diff_glmm1)#looks alright 
```


#### Post-Hoc Comparisons 
```{r}
#looks at herbivory within each management treatment just to see what differs 
emmeans_diff1<- emmeans(diff_glmm1, ~herbivory_trt|management_trt, adjust="tukey", df = 49)
pairs(emmeans_diff1, adjust="tukey")

#compare herbivory treatments to each other
emmeans_diff2<- emmeans(diff_glmm1,  ~pairwise~herbivory_trt, adjust="tukey", df = 49)
```

Table: GLMM Delta Richness Model
```{r}
#delta richness model summary
glmm_diff_sum<-summary(diff_glmm1)
#make into df and round up 
glmm_diff_df <- as.data.frame(glmm_diff_sum$coefficients$cond) %>% tibble::rownames_to_column("Term") %>% 
  mutate(across(c(Estimate, `Std. Error`, `z value`, `Pr(>|z|)`), ~round(.x, 3)))
glmm_diff_df

```

Table 2: ANova Delta Richness Model 
```{r}
#use anova table for results section 
#anova_diff

#convert to df, keep column names and round up
anova_diff_df<-as.data.frame(anova_diff) %>% rownames_to_column("Term") %>%  rename(Chisq = Chisq, Df= Df, p_value= `Pr(>Chisq)`)

#export to CSV 
#write.csv(anova_diff_df, "Table 2 Anova Delta Model.csv", row.names = FALSE)
```

Table: Post Hoc Comparisons Delta Richness 
```{r}
#extract data for post hoc comparisons tables 
#emeansdiff 2 = herbivory trt alone
#emeans diff 1 = m*h interction comparisons

#need to extract the contrasts from my pairs first
contrast_diff<-contrast(emmeans_diff1, method="pairwise", adjust= "tukey")
contrast_diff2<- contrast(emmeans_diff2, method="pairwise", adjust="tukey")

#herbivory and management group contrasts 
pairs_diff1_df<-as.data.frame(contrast_diff) %>% rename(Management_Treatment = management_trt, Contrast = contrast, Estimate = estimate, SE=SE, Df= df, t_ratio = t.ratio, p_value= p.value) %>% mutate(across(c(Estimate, SE, t_ratio), ~round(.x, 3)), p_value=round(p_value, 4))

#write.csv(pairs_diff1_df, "Table_S5_PostHoc_HerbMgmt_DeltaRichness.csv", row.names = FALSE)

#herbivory only contrasts for signif main effects
pairs_diff2_df<-as.data.frame(contrast_diff2) %>% rename( Contrast = contrast, Estimate = estimate, SE=SE, Df= df, t_ratio = t.ratio, p_value= p.value) %>% mutate(across(c(Estimate, SE, t_ratio), ~round(.x, 3)), p_value=round(p_value, 4))

#write.csv(pairs_diff2_df, "Table_S4_PostHoc_BetweenHerb_DeltaRich.csv", row.names = FALSE)


```

#### Fig. 3: Data Visualization Delta Richness Herbivory Treatment Main Effects
```{r}
# Plot the differences with error bars and custom colors for herbivory treatments

#use emmeans for estimated marginal means with 95% CI
emm_diff1 <-emmeans(diff_glmm1, ~herbivory_trt, df = 49)
emm_df2 <-as.data.frame(emm_diff1)

#uses tukey pairwise comparisons for our significant contrasts
diff_2324_contrasts<- as.data.frame(pairs(emm_diff1, adjust="tukey"))#get contrasts for our significance values from our emmans data 
#colors for plotting
herbivory_colors <- c("SB" = "#440154FF", "SX" = "#35B779FF","XX" = "#31688EFF")
management_colors <- c("control" = "#000000","fire" = "#CC79A7", "herbicide" = "#E69F00","mow"= "#0072B2")


#need our brackets redefined by y position for no overlap 
#need to be stacked
pval_diff1_df<- diff_2324_contrasts %>% filter(p.value <0.05) %>% # only get significant pairwise difference 
  mutate(group1 = str_split_fixed(contrast, " - ", 2)[,1], #this splits apart my contrasts SB - XX into groups
         group2= str_split_fixed(contrast, " - ",2)[,2],
         p.label=case_when( p.value<0.001 ~ "***", p.value<0.01~"**", p.value<0.05~"*"))  # assigns asterisks for pvalues 
# these are our factor levels and helps with x -axis bracket positions
herbivory_levels<-c("XX", "SX", "SB")
management_levels<- c("control", "fire", "herbicide", "mow")

#will use this data frame with ggpbur package to add to pvalues to ggplot
#keeps only significant pairwise comparisons
#contrast columns are split into group 1 and group 2 for brackets between consumer groups
pval_diff1_df<-pval_diff1_df %>% 
  mutate(group1=factor(group1, levels = herbivory_levels), #asociates groups with herbivory levels for sb,sx,xx
        group2=factor(group2, levels = herbivory_levels), 
      x.position1=as.numeric(group1), x.position2=as.numeric(group2))#should aligns each x position correctly

#need our brackets redefined by y position for no overlap 
pval_diff1_df<- pval_diff1_df  %>%  arrange(p.value) %>% #order for contrast significance
  mutate(contrast_index=row_number(), #this helps with stacking brackets 
        base_y= max (emm_df2$upper.CL)+2, #this will adjust our heightof brackets, may need to modify this
        y.position= base_y + (contrast_index -1) *0.7) %>%ungroup() #this staggers our brackets on the y
 
#plot both raw and estimatd means with 95% ci
ggplot()+
  # raw data points that are transparent
  geom_jitter (data=rich_diff, aes(x=herbivory_trt, y=diff_richness, color=herbivory_trt), size=2, alpha=0.3, shape=16, position = pos_jitter2)+ 
  # group means with CI bars from emmeans data 
  geom_point(data=emm_df2, aes(x=herbivory_trt,y=emmean, color=herbivory_trt ), position = pos_dodge, size=3)+
  #CI for error bars
  geom_errorbar(data = emm_df2, aes(x=herbivory_trt, ymin = lower.CL, ymax = upper.CL, color=herbivory_trt), width=0.2, position = pos_dodge)+
  #add line for zero to show difference 
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  scale_color_manual(values = herbivory_colors,labels= c( "XX"= "No Mammals", "SX"= "Small Mammals Only", "SB"= "Bison + Small Mammals ")) + # Use custom colors
  theme_classic2()+
  text_theme(base_size = 12)+ #theme we made earlier
  theme(legend.position = "none", plot.margin = margin(t=10, r=35, b=10, l=10))+ #should help with formating long text
  coord_cartesian(clip = "off")+ #helps for ggplot to not cut off words
  labs(x = "Consumer Treatment", y = "Change in Species Richness", color = "Consumer Treatment") +  # Update legend label
  scale_x_discrete(labels= c("XX" ="No Mammals", "SX"="Small Mammals", "SB"="Bison + Small Mammals"))+
  stat_pvalue_manual(data=pval_diff1_df, label = "p.label", xmin="x.position1", xmax = "x.position2", y.position = "y.position", tip.length = 0.01, bracket.size = 0.8, color = "black", size=6)
 
#save our figures
ggsave ("Fig3_DeltaSpecRIch.png",  width= 7.3, height=5, units = "in", dpi = 600) 
```


### Fig S2. Delta Richness Management X Herbivory Treatments 
```{r}
# Plot the differences with error bars and custom colors for herbivory treatments

#use emmeans for estimated marginal means with 95% CI
emm_diff2 <-emmeans(diff_glmm1, ~herbivory_trt|management_trt, df = 49)
emm_df3 <-as.data.frame(emm_diff2)
#uses tukey pairwise comparisons for our significant contrasts
diff_mgmt_contrasts<-as.data.frame(pairs(emm_diff2, adjust="tukey"))


#will use this data frame with ggpbur package to add to pvalues to ggplot 
pval_diff_mgmt<- diff_mgmt_contrasts %>% 
  filter(p.value <0.05) %>% # only get significant pairwise difference 
  mutate(group1 = str_split_fixed(contrast, " - ", 2)[,1], #this splits apart my contrasts SB - XX into groups
         group2= str_split_fixed(contrast, " - ",2)[,2],
         p.label=case_when( p.value<0.001 ~ "***", p.value<0.01~"**", p.value<0.05~"*"))  # assigns asterisks for pvalues 

# these are our factor levels and helps with x -axis bracket positions
herbivory_levels<-c("SB", "SX", "XX")
management_levels<- c("control", "fire", "herbicide", "mow")

#need our brackets redefined by y position for no overlap 
#need to be stacked
pval_diff_mgmt<-pval_diff_mgmt %>% 
  mutate(group1=factor(group1, levels = herbivory_levels), #asociates groups with herbivory levels for sb,sx,xx
        group2=factor(group2, levels = herbivory_levels), 
        management_trt=factor(management_trt, levels = management_levels), 
        xnum=as.numeric(management_trt), # calculate the x position for each treatment
        dodge_width=0.5, # i used this width for my dodge in my ggplot and we want it replicated
        x.position1=xnum-dodge_width/2+(as.numeric(group1)-1)* (dodge_width/2), #aligns each x position correctly
        x.position2=xnum-dodge_width/2+(as.numeric(group2)-1)* (dodge_width/2),)

#need our brackets redefined by y position for no overlap 
pval_diff_mgmt <- pval_diff_mgmt %>% group_by(management_trt) %>%  arrange(p.value) %>% #order for contrast significance
  mutate(contrast_index=row_number(), #this helps with stacking brackets 
        base_y= max (emm_df3$upper.CL)+0.5, #this will adjust our heightof brackets, may need to modify this
        y.position= base_y + (contrast_index -1) *0.5) %>% ungroup()#this staggers our brackets on the y


#plot both raw and estimated means with 95% ci
ggplot()+
  # raw data points that are transparent
  geom_jitter (data=rich_diff, aes(x=management_trt, y=diff_richness, color=herbivory_trt), size=2, alpha=0.3, shape=16, position = pos_jitter2)+ 
  # group means with CI bars from emmeans data 
  geom_point(data=emm_df3, aes(x=management_trt,y=emmean, color=herbivory_trt ), position = pos_dodge, size=3)+
  #CI for error bars
  geom_errorbar(data = emm_df3, aes(x=management_trt, ymin = lower.CL, ymax = upper.CL, color=herbivory_trt), width=0.2, position = pos_dodge)+
  #add line for zero to show difference 
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  scale_color_manual(values = herbivory_colors,labels= c( "XX"= "No Mammals", "SX"= "Small Mammals ", "SB"= "Bison + Small Mammals")) + # Use custom colors
  theme_classic2()+
  text_theme(base_size = 12)+ #our theme
    theme(legend.position = "bottom", plot.margin = margin(t=10, r=35, b=10, l=10))+ #need this for long text
  coord_cartesian(clip = "off")+ #helps to not cut off text when saving plots 
  labs(x = "Management Treatment", y = "Difference in Species Richness",color = "Consumer Treatment") +  # Update legend label
 scale_x_discrete(labels = c("control"="Control", "fire"="Fire", "mow"="Mow", "herbicide"="Herbicide"))+
  stat_pvalue_manual(data=pval_diff_mgmt, label = "p.label", xmin="x.position1", xmax = "x.position2", y.position = "y.position", tip.length = 0.01, bracket.size = 0.8, color = "black", size = 6)
  
#save file
  ggsave ("Fig S2_DeltaRichMgmt.png",  width= 7.3, height=5, units = "in", dpi = 600) 
  
```

## Shannon and Evenness Metrics


#### Data Cleaning and Set up 
```{r,results='hide', fig.show='hide'}

cover <- cover_clean
str(cover)
cover4<-cover[, -c(11:15)] #remove extra columns with taxon names and etc for now 

cover4$block<- as.factor(cover4$block) # change block/plot to  factor for data wrangling
cover4$plot<- as.factor(cover4$plot)  
```

```{r,results='hide', fig.show='hide'}
cover4<- cover4 %>%  
  filter(live==1) %>%  # remove non plant 
  filter(unknown==0) %>%  # remove unknowns
  mutate(unique_id=paste(year,block, plot, management_trt, herbivory_trt, bison_pasture, sep = "_")) %>%  # make a unique ID column because we have multiple plot #s over the years but each is unique and will help with merging data later 
  relocate(unique_id, .before = 1) %>%  # needs to be the first column
  filter(code != "" & !is.na(code)) %>%  #remove rows with empty plant code 
  mutate(cover= as.numeric(cover)) %>% 
  mutate(cover=replace_na(cover, 0)) # NAs need to be zero values to help when we flip to wide format 
  
cover4$herbivory_trt<-factor(cover4$herbivory_trt, levels=c("XX", "SX", "SB"))

 
# need wide matrix format 
 div_wide<- cover4 %>%  pivot_wider(names_from = code, values_from = cover, values_fn = (cover=sum), values_fill = list(cover=0)) # long data to wide for matrix, use the values to fil to remove nas and replace with zeros 
 
```

```{r,results='hide', fig.show='hide'}

#make sure plot and year are factors for merging 
graze_Aug$plot<-as.factor(graze_Aug$plot)
graze_Aug$year<-as.factor(graze_Aug$year)
div_wide$plot<-as.factor(div_wide$plot)
div_wide$year<-as.factor(div_wide$year)

# join both into new df
div_wide2<- div_wide %>%  left_join(graze_Aug %>% dplyr:: select(plot, year, Weighted_Grazing_Influence), by=c("plot", "year"))

#fill in missing values, graze inlfuence was done with block/year so some plots will be missing data
div_wide2 <-div_wide2 %>% group_by(block, year) %>% mutate(block_graze_value =first(Weighted_Grazing_Influence[!is.na(Weighted_Grazing_Influence)&Weighted_Grazing_Influence !=0]),Weighted_Grazing_Influence=ifelse(is.na(Weighted_Grazing_Influence)|Weighted_Grazing_Influence==0,block_graze_value, Weighted_Grazing_Influence)) %>% ungroup() %>% dplyr::select(-block_graze_value)

# replae any truly missing values with zero. this represents actual plots/years with no recorded bison influence
div_wide2<-div_wide2 %>% mutate(Weighted_Grazing_Influence=ifelse(is.na(Weighted_Grazing_Influence), 0, Weighted_Grazing_Influence))

```


Calculate shannon and evennes metrics
```{r,results='hide', fig.show='hide'}
#shannon diveristy using div_wide to species only matrix df
#separate environmental and treatments first from abundance matrix
#what is left wil be species only matrix with abundance data plus the unique id column
 div_env_df<-div_wide2[,-c(10:71)] #remove unneded columns
 species_matrix <-div_wide2[,-c(2:9,72)] #keep unique id column for merging later

 #calculate shannon, richness and both evenness metrics  
 #remove the first column before calc. 
shannon=diversity(species_matrix[,-c(1)], index = "shannon") 
richness =specnumber(species_matrix[,-c(1)])  
pielou_E =ifelse(richness>1, shannon/log(richness), NA) 

#combine environmental with our diversity metrics data 
div_metrics<-div_env_df %>% dplyr::select(unique_id, year, block, plot, management_trt, herbivory_trt, bison_pasture, Weighted_Grazing_Influence) %>% mutate(shannon=shannon, richness=richness, pielou_E=pielou_E)

#calculate other Evennes metric (PER Avoilio et al 2019 methods*)
calc_evar<-function(cover) {
  cover<-cover[cover>0] #remove species with zero abundnace
  if(length(cover)<2)return(NA) #evar needs at least two species to calculate variance in abundance
  var_ln<-var(log(cover)) #calc variance of log transformed abundnance
  evar<-1-(2/pi)* atan(var_ln) #calc our evar metric 
  return(evar)}
#apply evar function to each row
evar_values<-apply(species_matrix[,-c(1)],1, calc_evar)
#add evar values to our diversity metrics df
div_metrics$evar<-evar_values
```


Shannon Diversity 2024
```{r,results='hide', fig.show='hide'}
#filter pasture and year
shannon_W24<- div_metrics %>% filter(year==2024, bison_pasture=="yes")
#make sure all are factors for analysis
shannon_W24$herbivory_trt<-as.factor(shannon_W24$herbivory_trt)
shannon_W24$management_trt<-as.factor(shannon_W24$management_trt)

```

```{r,results='hide', fig.show='hide'}
#hisorgram 
ggplot(shannon_W24, aes(x= shannon))+
  geom_histogram()+
  theme_bw()+
 labs(title = "Distribution of shannon Data", x= "shannon", y= "Frequency")
#has a slight left tail but will check with diagnostics
#but will not use gamma even though cont and positive, i do not see a strong right skew
```

#### Fit model for Shannon diversrity 
```{r,results='hide', fig.show='hide'}
# model 
shan_glmm1<-glmmTMB(shannon~herbivory_trt*management_trt+(1|block/plot), data=shannon_W24, family = gaussian )
summary(shan_glmm1)
#anova 
shan_anov<-Anova(shan_glmm1, type="III")
#significant interaction
shan_anov        
#weighted grazing influence is not significant and removed from model
#check assumptions

sim_res_shan1<-simulate_residuals(shan_glmm1)
plot(sim_res_shan1)
check_model(shan_glmm1)
#looks fine  no issues with normality, qq plots good, posterior overlap is good, and no other red flags 
#littler curviature in residuals vs fitted but not big deal 


```



Model summary: Summary of GLMM for shannon diversity  in response to management, herbivory and their interaction using data from 2024. 
```{r,results='hide', fig.show='hide'}

glmm_shan_sum<-summary(shan_glmm1)

#make into df and round up 
glmm_shan_df <- as.data.frame(glmm_shan_sum$coefficients$cond) %>% tibble::rownames_to_column("Term") %>% 
  mutate(across(c(Estimate, `Std. Error`, `z value`, `Pr(>|z|)`), ~round(.x, 3)))

#model summary
glmm_shann_df

```


Table S2: Type III Wald Chisquare tests
```{r,results='hide', fig.show='hide'}
#use anova table for results section 
#shan_anov
#convert to df, keep column names and round up
anova_shan_df<-as.data.frame(shan_anov) %>% rownames_to_column("Term") %>% rename(Chisq = Chisq, Df= Df, p_value= `Pr(>Chisq)`)

#export to CSV 
#write.csv(anova_shan_df, "Table S2_Shannon_Anova.csv", row.names = FALSE)
```




Pielou evenness 
```{r,results='hide', fig.show='hide'}
evenness_W24<-div_metrics %>% filter(year==2024, bison_pasture=="yes")

even_glmm2<-glmmTMB(pielou_E ~herbivory_trt*management_trt+(1|block/plot), data=evenness_W24, family=gaussian)
summary(even_glmm2)
pielou_anova<-Anova(even_glmm2)
#no significance interactions
#one treatment (mow) within the glmm  coefficient table was but overall ANOVA tests output shows no significant main or interactive effects 
```

```{r,results='hide', fig.show='hide'}
sim_pielou_g<- simulateResiduals(even_glmm2)
plot(sim_pielou_g)
check_model(even_glmm2)
testDispersion(sim_pielou_g)
testZeroInflation(sim_pielou_g)

#not bad , no signif KS, disperions or outliers
#residual vs predicted was much better

```


```{r,results='hide', fig.show='hide'}
pielou_beta<-glmmTMB(pielou_E ~herbivory_trt*management_trt+ (1|block/plot), data=evenness_W24, family = beta_family(link = "logit"))
summary(pielou_beta)

sim_pielou_b<- simulateResiduals(pielou_beta)
plot(sim_pielou_b)
check_model(pielou_beta)
testDispersion(sim_pielou_b)
testZeroInflation(sim_pielou_b)


```

Same as above. Also beta model reinforces that the single coefficient is non-significant here. Will still use guassian model for results


Table S3: Summary of anova for pielou's evennes in response to management, herbivory and their interaction using data from 2024. 
```{r,results='hide', fig.show='hide'}
#convert to df, keep column names and round up
anova_pielou_df<-as.data.frame(pielou_anova) %>% rownames_to_column("Term") %>% rename(Chisq = Chisq, Df= Df, p_value= `Pr(>Chisq)`)

#export to CSV 
#write.csv(anova_pielou_df, "TableS3_Pielou_Anova.csv", row.names = FALSE)

```




## Comm Comp 2024

### Data Cleaning and Set up for Matrix
```{r}
cover2<-cover[, -c(11:15)] #remove extra columns with taxon names and etc for now 

cover2$block<- as.factor(cover2$block) # change block/plot to  factor for data wrangling
cover2$plot<- as.factor(cover2$plot)  
```

```{r}
cover3<- cover2 %>%  
  filter(live==1, unknown==0) %>%  # remove unknowns and non plants 
  mutate(unique_id=paste(year,block, plot, management_trt, herbivory_trt, bison_pasture, sep = "_")) %>%  # make a unique ID column because we have multiple plot #s over the years but each is unique and will help with merging data later 
  relocate(unique_id, .before = 1) %>%  # needs to be the first column
  filter(code != "" & !is.na(code)) %>%  #remove rows with empty plant code 
  mutate(cover= as.numeric(cover)) %>% #we need cover value column to be numeric
  mutate(cover=replace_na(cover, 0)) # NAs need to be zero values to help when we flip to wide format 
  
  
# need matrix format 
 df_wide<- cover3 %>%  pivot_wider(names_from = code, values_from = cover, values_fn = (cover=sum), values_fill = list(cover=0)) # long data to wide for matrix, use the values to fil to remove nas and replace with zeros 


```

```{r}
## double check data for outliers or anything funky before moving forward
#cover3 %>% 
 # group_by(herbivory_trt, management_trt, year) %>% 
 # summarize(
 #   n              = n(),
  #  n_cover_NA     = sum(is.na(cover)),
  #  cover_min      = min(cover, na.rm = TRUE),
  #  cover_max      = max(cover, na.rm = TRUE),
  #  mean_cover     = mean(cover, na.rm = TRUE),
  #  n_cover_negative = sum(cover < 0, na.rm = TRUE),
  #  .groups = "drop"
 # ) %>% 
 # arrange(herbivory_trt, management_trt, year) %>% print(n=41)


#found one extreme outlier, one comma were there shoudl have been . all NAs are removed and we have no negative values

```

```{r}
#step 1
# subset data for only west pasture 
df_W24<- df_wide %>% filter(bison_pasture== "yes") %>% filter(year=="2024")

#need a new data frame with our bison raster data 
graze_Aug$plot<-as.factor(graze_Aug$plot) #everything needs to be the same to join
graze_Aug$year<-as.factor(graze_Aug$year)
df_W24$year <-as.factor(df_W24$year)
df_W24$block<-as.factor(df_W24$block)
df_W24$plot<-as.factor(df_W24$plot)

# join both into new df
df_W24b<- df_W24 %>% left_join(graze_Aug %>% dplyr:: select(plot, year, Weighted_Grazing_Influence), by=c("plot", "year"))

#fill in missing values since grazing influene was esitmated at block/year level associated with bison plots
#need all plots filled in with data by blcok 
df_W24b <-df_W24b %>% group_by(block, year) %>% 
  mutate(block_graze_value =first(Weighted_Grazing_Influence[!is.na(Weighted_Grazing_Influence)&Weighted_Grazing_Influence !=0]), Weighted_Grazing_Influence=ifelse(is.na(Weighted_Grazing_Influence)|Weighted_Grazing_Influence==0, block_graze_value, Weighted_Grazing_Influence)) %>% ungroup() %>% dplyr::select(-block_graze_value)

# replace any missing values with zero as these are plots/years that have no actuall recorded bison influence data
df_W24b<-df_W24b %>% mutate(Weighted_Grazing_Influence=ifelse(is.na(Weighted_Grazing_Influence), 0, Weighted_Grazing_Influence))

```

```{r}
# and remove columns to create cover df and environmental df 
cover_W24<- df_W24b[, -c(2:9)] # keep unique ID for merging purposes later on

env_W24 <- df_W24b[, -c(7:83)] # may need to drop year/block/plot if using envfit (will show up on plots)

```

```{r}
# step 2: standardize the community matrix using site total with decostand

covTotal_W24 <- decostand(cover_W24 [, -c(1)], "total") # dividies by margin total and removes first column 

```

```{r, results='hide', fig.show='hide'}
 # step 3: check dimensions

dimcheckMDS(covTotal_W24, distance = "bray", k=6, trymax = 50, autotransform = FALSE)
#stress test for 2 is great (0.168)
```

```{r, results='hide'}
#step 4: run metaMDS 

Tot_NMDS_W24<- metaMDS(covTotal_W24, distance = "bray", k=2, trymax = 500, maxit=500, autotransform = FALSE, expand = FALSE)

#best solution repeated 2 time 

```


```{r}
#step 5: standardize the data for dissimilarity distances using the data that was standardized by site total above with decostand 

cov_bray_dis_W24<-vegdist(covTotal_W24, "bray")

# make sure original data frame (subset matrix with env) treatments are all factors 
df_W24b$management_trt<-as.factor(df_W24b$management_trt)
df_W24b$herbivory_trt<- as.factor(df_W24b$herbivory_trt)

```

Permanova
```{r, results='hide'}
#step 6: run permanova with adonis2
permW24<-adonis2(cov_bray_dis_W24~management_trt*herbivory_trt+Weighted_Grazing_Influence, permutations = 999, data = df_W24b, strata = df_W24b$block, by="terms")
permW24
#significant interaction and bison covariate
```



```{r}
#build an interaction term to see if there are significant disperions between all the groupings
df_W24b$inter_group<-interaction(df_W24b$management_trt, df_W24b$herbivory_trt,  sep="_")

perm_24_inter<-betadisper(cov_bray_dis_W24, group = df_W24b$inter_group)
full_permdisp_results<-permutest(perm_24_inter,  permutations = 999)

#this was significant- we will reanalyze this on our subset data 
```


Table: PERMANOVA 2024
```{r}
# extract permanova results for table and write as csv

permw24_df<- as.data.frame(permW24) %>% rownames_to_column("Term") %>% 
  rename(Df=Df, Sum_of_Sqs = SumOfSqs, R2 = R2, F=F, p_value= `Pr(>F)`) 

#write.csv(permw24_df, "Table 5 Permanova 2024.csv", row.names = FALSE)
```

Table permdisp test
```{r}

# extract permanova results for table and write as csv

full_permdisp_results_df<- as.data.frame(full_permdisp_results$tab ) %>% rownames_to_column("Term") %>%  rename( p_value= `Pr(>F)`) 

#write.csv(full_permdisp_results_df, "Table BEtaDispersion_2024all.csv", row.names = FALSE)

```


Post Hoc Comparisons 
```{r, results='hide'}
pairsW24_h<-pairwise.adonis2(cov_bray_dis_W24 ~herbivory_trt, data = df_W24b, permutations =999, p.adjust.methods ="holm")
pairsW24_h
#herbivory comparisons without interactions (Sb vs SX, p= 0.087 and SB vs XX p= 0.102, SX and XX p= 0.281) 

pairsW24_m<-pairwise.adonis2(cov_bray_dis_W24 ~management_trt, data = df_W24b, permutations =999, p.adjust.methods ="holm")
pairsW24_m
#management comparisons alone without herbivory * all plant communities are different 
#Fire is significantly different from mow, herbcide, and control (p< 0.01)
# mow is different from fire, hferbcide and control (p<0.01) 
# herbicide and control are different (p=0.001) 


#create interaction grouping variable-which we already have one "inter_group" 

pairsW24_inter<-pairwise.adonis2(cov_bray_dis_W24 ~inter_group, data = df_W24b, permutations =999, p.adjust.methods ="holm")
pairsW24_inter
 # 

```


Table . Pairwise comparisons of plant community composition for interactive  effects in west pasture 2024. 
```{r}
pair_names<-setdiff(names(pairsW24_inter), "parent_call") #get rid of the parent call 

#get each pairwise table and add a column identfying where it came from since each is stored separately
pairwise_df_inter<-lapply(pair_names, function(name){ # each one of our comparisosn has its own table we need to extract 
  result<-pairsW24_inter[[name]]
  result$Comparison<-name # this will have our comparison names 
  return(result)}) %>% bind_rows() # this should combine into one "df" 

#add significant column for asterisks and keep only rows needed for our final table
# !is.na filters any non-model rows with no f statistic
pairwise_df_inter_csv <-pairwise_df_inter %>% mutate( sig = case_when (`Pr(>F)`<=0.001~ "***", `Pr(>F)`<=0.01~ "**", `Pr(>F)`<=0.05~ "*", `Pr(>F)`<=0.1~ ".",TRUE ~ "")) %>% 
#keep only the model rows with the important information
filter(!is.na(F)) %>% dplyr::select(Comparison, Df, SumOfSqs, R2, F, `Pr(>F)`, sig) 

#write.csv(pairwise_df_inter_csv, "Table Pairwise Permanova Interactions.csv", row.names = FALSE)
```

Table Comparisons Comm Comp for Herbivory Main Effects 2024
```{r}
#same as above


pair_names_h<-setdiff(names(pairsW24_h), "parent_call") #get rid of the parent call 

pairwise_df_h<-lapply(pair_names_h, function(name){ # each one of our comparisosn has its own table we need to extract 
  result<-pairsW24_h[[name]]
  result$Comparison<-name # this wil have our comparison names 
  return(result)}) %>% bind_rows() # this should combine into one "df" 

#add significant column for asterisks 
pairwise_df_h_csv <-pairwise_df_h %>% mutate( sig = case_when ( `Pr(>F)`<=0.001~ "***", `Pr(>F)`<=0.01~ "**", `Pr(>F)`<=0.05~ "*", `Pr(>F)`<=0.1~ ".",TRUE ~ "")) %>% 
#keep only the model rows with the important information
filter(!is.na(F)) %>% dplyr::select(Comparison, Df, SumOfSqs, R2, F, `Pr(>F)`, sig) 

#write.csv(pairwise_df_h_csv, "Table Pairwise Permanova Herbivory.csv", row.names = FALSE)

```

Table Comparisons Management main effects Permanova 2024
```{r}

#same as above


pair_names_m<-setdiff(names(pairsW24_m), "parent_call") #get rid of the parent call 

pairwise_df_m<-lapply(pair_names_m, function(name){ # each one of our comparisosn has its own table we need to extract 
  result<-pairsW24_m[[name]]
  result$Comparison<-name # this wil have our comparison names 
  return(result)}) %>% bind_rows() # this should combine into one "df" 

#add significant column for asterisks 
pairwise_df_m_csv <-pairwise_df_m %>% mutate( sig = case_when (`Pr(>F)`<=0.001~ "***", `Pr(>F)`<=0.01~ "**", `Pr(>F)`<=0.05~ "*", `Pr(>F)`<=0.1~ ".",TRUE ~ "")) %>% 
#keep only the model rows with the important information
filter(!is.na(F)) %>% dplyr::select(Comparison, Df, SumOfSqs, R2, F, `Pr(>F)`, sig) 

#write.csv(pairwise_df_m_csv, "Table Pairwise Permanova Management.csv", row.names = FALSE)

```



#### Data Visualization 2024 West Pasture 

```{r}


#extract scores from nmds using site scores
nmds_W24<- as.data.frame(scores(Tot_NMDS_W24, display = "sites"))

#add treatment factors from our env df to our nmds scores
nmds_W24$management_trt <- env_W24$management_trt
nmds_W24$herbivory_trt <- env_W24$herbivory_trt

##  colors and labels 
management_colors <- c("control" = "#000000","fire" = "#CC79A7", "herbicide" = "#E69F00","mow"= "#0072B2")
herbivory_colors <- c("SB" = "#440154FF", "SX" = "#35B779FF", "XX" = "#31688EFF")
facet_labels<-c("control"="Control", "fire"="Fire", "mow"="Mow", "herbicide"="Herbicide")


#calculate centroids and SE for each group then we mutliply by 1.96 to get our 95% C.I. 
cent_W24<- nmds_W24 %>% group_by(management_trt, herbivory_trt) %>% 
  summarize(NMDS1_mean =mean(NMDS1), 
            NMDS2_mean = mean(NMDS2), 
            NMDS1_se=sd(NMDS1)/sqrt(n()), 
            NMDS2_se=sd(NMDS2)/sqrt(n()), 
            .groups = "drop")

#plot with raw and mean/centroids data 
ggplot()+
  #raw data points 
  geom_point(data = nmds_W24, aes(x=NMDS1, y=NMDS2, color=herbivory_trt),shape=18,  size=3, alpha=0.3, position=position_jitter(width = 0.02, height=0.02))+
  facet_wrap(~management_trt, labeller=as_labeller(facet_labels))+
  #reorder for error bars first so centroids "sit" on top
  geom_errorbar(data=cent_W24, aes(x= NMDS1_mean, ymin=NMDS2_mean - 1.96*NMDS2_se, ymax= NMDS2_mean + 1.96*NMDS2_se, color=herbivory_trt), width=0.05, alpha=0.8, linewidth=0.4)+
  geom_errorbarh(data=cent_W24, aes(y=NMDS2_mean, xmin =NMDS1_mean - 1.96*NMDS1_se, xmax= NMDS1_mean+ 1.96*NMDS1_se, color=herbivory_trt), height=0.05, alpha=0.8, linewidth=0.4)+
  #centroids
  geom_point(data=cent_W24, aes(x= NMDS1_mean, y=NMDS2_mean, color=herbivory_trt), size=4, alpha=1)+
  scale_color_manual(values=herbivory_colors, labels= c("SB"= "Bison + Small Mammals", "SX"= "Small Mammals", "XX"= "No Mammals") )+
  labs( x = "NMDS 1", y= "NMDS 2", color= "Consumer Treatment")+
  theme_minimal()+
  theme(legend.position = "bottom", plot.margin = margin(t=10, r=35, b=10, l=10))+
  coord_cartesian(clip = "off")+ #keeps from cutting off text
  text_theme(base_size = 12)
  

ggsave("Ch2_Figure5NMDS.png", width=7.3, height=5.5,units= "in", dpi=600)


```

#### Subset Data to run betadispersion tests within each management treatment

Control Subset
```{r, results='hide'}
# subset for control treatment

#first add back the unique id column to help filter 

covTotal_w24_2<-cbind(unique_id = cover_W24[[1]], covTotal_W24)

# subset env data 
env_cont <-env_W24 %>% filter(management_trt=="control")
#subset community matrix 
cov_cont <-covTotal_w24_2 %>% filter( unique_id %in% env_cont$unique_id)



 #rerun vegdist and adonis2 on control only data 

dist_cont <-vegdist(cov_cont[,-c(1)], method = "bray")

env_cont$herbivory_trt<-as.factor(env_cont$herbivory_trt)

perm_cont<-adonis2(dist_cont ~ herbivory_trt +Weighted_Grazing_Influence,permutations = 999, data = env_cont, strata = env_cont$block, by = "terms")
perm_cont

pairwise.adonis2(dist_cont ~herbivory_trt , data = env_cont, permutations =999, p.adjust.methods ="holm")

beta_24c <- betadisper(dist_cont, group = env_cont$herbivory_trt)
permutest(beta_24c)

#
#0.202-permanova
#post hoc no differences (p>0.05)
#betadisp = 0.172
```

Mow Subset
```{r,results='hide', fig.show='hide'}
# subset for mow treatment

# subset env data 
env_mow <-env_W24 %>% filter(management_trt=="mow")
#subset community matrix 
cov_mow <-covTotal_w24_2 %>% filter( unique_id %in% env_mow$unique_id)

 #rerun vegdist and adonis2 on mow only data 

dist_mow <-vegdist(cov_mow[,-c(1)], method = "bray")

env_mow$herbivory_trt<-as.factor(env_mow$herbivory_trt)

perm_mow<-adonis2(dist_mow ~ herbivory_trt +Weighted_Grazing_Influence,permutations = 999, data = env_mow, strata = env_mow$block, by = "terms")
perm_mow

pairwise.adonis2(dist_mow ~herbivory_trt, data = env_mow, permutations =999, p.adjust.methods ="holm")

beta_24m <- betadisper(dist_mow, group = env_mow$herbivory_trt)
permutest(beta_24m)

#p=0.842
#post hoc p>0.05
#permdisp p>0.05

```



```{r,results='hide', fig.show='hide'}
# subset for fire treatment

# subset env data 
env_fire <-env_W24 %>% filter(management_trt=="fire")
#subset community matrix 
cov_fire <-covTotal_w24_2 %>% filter( unique_id %in% env_fire$unique_id)


 #rerun vegdist and adonis2 on fire only data 

dist_fire <-vegdist(cov_fire[,-c(1)], method = "bray")

env_fire$herbivory_trt<-as.factor(env_fire$herbivory_trt)

perm_fire<-adonis2(dist_fire ~ herbivory_trt+Weighted_Grazing_Influence,permutations = 999, data = env_fire, strata = env_fire$block, by = "terms")
perm_fire

pairwise.adonis2(dist_fire ~herbivory_trt, data = env_fire, permutations =999, p.adjust.methods ="holm")

beta_24f <- betadisper(dist_fire, group = env_fire$herbivory_trt)
permutest(beta_24f)

# p=0.191
#post hoc SB vs SX p =0.084
#permdisp p>0.05
```



```{r, results='hide'}
# subset for herbicide treatment

# subset env data 
env_herb <-env_W24 %>% filter(management_trt=="herbicide")
#subset community matrix 
cov_herb <-covTotal_w24_2 %>% filter( unique_id %in% env_herb$unique_id)

 #rerun vegdist and adonis2 on herbicide only data 

dist_herb <-vegdist(cov_herb[,-c(1)], method = "bray")

env_herb$herbivory_trt<-as.factor(env_herb$herbivory_trt)

perm_herb<-adonis2(dist_herb ~ herbivory_trt+Weighted_Grazing_Influence,permutations = 999, data = env_herb, strata = env_herb$block, by = "terms")
perm_herb

herb_pairs<-pairwise.adonis2(dist_herb ~herbivory_trt, data = env_herb, permutations =999, p.adjust.methods ="holm")

beta_24h <- betadisper(dist_herb, group =  env_herb$herbivory_trt)
permutest(beta_24h, pairwise = TRUE)


#pairwise test showed that SB plots significantly differ from SX and XX in dispersion

#p=0.001
#permdisp p=0.006
#post hoc all differ 
```


 
```{r,results='hide', fig.show='hide'}
#get distances from each plot to its group centorid from our beta dispersion analysis 
h_distances<-beta_24h$group.distances
plot(h_distances)#quick plot to visualize-not as helpful 
h_distances #print distances
boxplot(beta_24h) #better visualization 
```

#### Data Visualization Herbicide Distance to Centroid Plot
```{r}
#custom colors
herbivory_colors <- c("SB" = "#440154FF", "SX" = "#35B779FF", "XX" = "#31688EFF")

#need df fro our ggplot using distances of each plot from centroid
distance_df<- data.frame(distance = beta_24h$distances, group= beta_24h$group) %>% 
  mutate(group=factor(group, levels= c ("XX", "SX", "SB")))
#we want both plots and the boxplot to summarize distribution
#jittered are individual plots 
ggplot( distance_df, aes(x= group, y= distance, color=group, fill=group))+
  geom_boxplot(alpha=0.25, outlier.shape = NA, width=0.6)+
  geom_jitter(width=0.12, size=2, alpha=0.6)+ #individual plots 
  stat_summary(fun=mean, geom="point", size=3, shape=18)+
  theme_classic2()+
  text_theme(base_size = 12)+
   scale_color_manual(values=herbivory_colors )+
 scale_fill_manual(values=herbivory_colors )+
  scale_x_discrete(labels= c( "XX"= "No Mammals", "SX"= "Small Mammals","SB"= "Bison + Small Mammals"))+
  labs( x = "Consumer Treatment", y= "Distance to Centroid(multivariate dispersion)", color= "Consumer Treatment", fill="Consumer Treatment")+
  theme(legend.position = "none", plot.margin = margin(t=10, r=35, b=10, l=10))+
  coord_cartesian(clip = "off")

ggsave("Ch2_SI_Herb_Distance.png",width=7.3, height=6.5,units= "in", dpi=600)

```

Table: PERMANOVA Subset Herbicide
```{r}
# extract permanova results for table and write as csv

perm_herb_df<- as.data.frame(perm_herb) %>% rownames_to_column("Term") %>% 
  rename(Df=Df, Sum_of_Sqs = SumOfSqs, R2 = R2, F=F, p_value= `Pr(>F)`) 

#write.csv(perm_herb_df, "Table Herb Permanova.csv", row.names = FALSE)
```

Table Comparisons Comm Comp for Herbicide Subset
```{r}
#same as above notes

pair_names_herb<-setdiff(names(herb_pairs), "parent_call") #get rid of the parent call 

pairwise_df_herb<-lapply(pair_names_herb, function(name){ # each one of our comparisosn has its own table we need to extract 
  result<-herb_pairs[[name]]
  result$Comparison<-name # this wil have our comparison names 
  return(result)
}) %>% bind_rows() # this should combine into one "df" 

#add significant column for asterisks 
pairwise_df_herb_csv <-pairwise_df_herb %>% mutate( sig = case_when (
   `Pr(>F)`<=0.001~ "***", `Pr(>F)`<=0.01~ "**", `Pr(>F)`<=0.05~ "*", `Pr(>F)`<=0.1~ ".",TRUE ~ "")) %>% 
#keep only the model rows with the important information
filter(!is.na(F)) %>% dplyr::select(Comparison, Df, SumOfSqs, R2, F, `Pr(>F)`, sig) 

#write.csv(pairwise_df_herb_csv, "Table Pairwise Subset Herbicide.csv", row.names = FALSE)

```

Table: Beta Dispersion for Herbicide Treatment 
```{r,results='hide', fig.show='hide'}
# rerun perm test
bd_test <- permutest(beta_24h, pairwise = TRUE)
bd_test
# Extract the ANOVA- table data 
bd_df <- as.data.frame(bd_test$tab) %>%   # $tab holds the ANOVA-style table
rownames_to_column(var = "Term") %>% 
  rename(Sum_of_Sqs = `Sum Sq`, Mean_Sq= `Mean Sq`, p_value= `Pr(>F)`) 

#write.csv(bd_df, "Table S Herbicide_BetaDisp2024.csv", row.names = FALSE)


```



## 2023 & 2024 NMDS 

#### data wrangling

```{r,results='hide', fig.show='hide'}
 

df_34b<-df_wide %>% 
  filter(year %in% c("2023", "2024")) %>%  
  filter(bison_pasture=="yes") %>% 
  filter(rowSums(.[,-c(1:9)])>0) # keep only rows that species/code columns are more than zero, rows with only zeros are filtered out

#need a new data frame with our bison raster data 
graze_Aug$plot<-as.factor(graze_Aug$plot) #everything needs to be the same to join
graze_Aug$year<-as.factor(graze_Aug$year)
df_34b$year <-as.factor(df_34b$year)
df_34b$block<-as.factor(df_34b$block)
df_34b$plot<-as.factor(df_34b$plot)

# join both into new df
df_34b<- df_34b %>% 
  left_join(graze_Aug %>% 
             dplyr:: select(plot, year, Weighted_Grazing_Influence), by=c("plot", "year"))

#fill in  missing values since bison influence was on block/year and we want all plots to have same data 
df_34b <-df_34b %>% group_by(block, year) %>% 
  mutate(block_graze_value =first(Weighted_Grazing_Influence[!is.na(Weighted_Grazing_Influence)&Weighted_Grazing_Influence !=0]),Weighted_Grazing_Influence=ifelse(is.na(Weighted_Grazing_Influence)|Weighted_Grazing_Influence==0,   block_graze_value,Weighted_Grazing_Influence)) %>% ungroup() %>% dplyr::select(-block_graze_value)

# fill in zero if it is missing now it is truly a zero for grazing influence
df_34b<-df_34b %>% mutate(Weighted_Grazing_Influence=ifelse(is.na(Weighted_Grazing_Influence), 0, Weighted_Grazing_Influence))



```

 
```{r,results='hide', fig.show='hide'}
#remove unecessary columns in data to create cover and separate environmental data frames

comm_34b <-df_34b[, -c(2:9,84)] # keep first column with unique ID to help merging later

env_34b<-df_34b[, -c(10:83)] #removes all plant code/cover columns keeping only env data

#standardize community matrix using site total with decostand
covTot34b<-decostand(comm_34b [,-c(1)], "total") # divided cover values by margin total and removes first column unique ID 


```

```{r, results='hide', fig.show='hide'}
# check the numbe rof dimension we need for the NMDS using Bray-Curtis dissmiliarity
dimcheckMDS(covTot34b, distance = "bray", k=6, trymax = 100, autotransform = FALSE)

#K=2 0.176

```

```{r, results='hide'}
#run metamds 
Tot_NMDS34b<-metaMDS(covTot34b, k=2, distance = "bray", trymax = 1000, autotransform = FALSE, expand = FALSE)
#best solution was repeated 2 times
```

```{r, results='hide'}
#need to standardize the data for dissmiliarity distances using the data that was standardized by site total above (with decostand=covTot...)
cov_bray_dis34b<-vegdist(covTot34b, "bray")

#need the data frame that has all the cover matrix plus env columns (original df_...)
df_34b$management_trt<-as.factor(df_34b$management_trt) #make sure all treatments are factors
df_34b$year<-as.factor(df_34b$year)
df_34b$herbivory_trt<-as.factor(df_34b$herbivory_trt)



perm2324<-adonis2(cov_bray_dis34b~management_trt*herbivory_trt*year+Weighted_Grazing_Influence, permutations = 999, data = df_34b,strata = df_34b$block, by="terms")


#using by terms for an anova like partitioing of total variation, tests each term while holding others constant
perm2324


#no 3 way interaction
#management and  herbiovry is nearly significant p=0.054
#mgmt * year : p=0.001
#all main effects <= 0.01



```


Table: PERMANOVA
```{r,results='hide', fig.show='hide'}
# extract permanova results for table and write as csv

perm2324_df<- as.data.frame(perm2324) %>% rownames_to_column("Term") %>% 
  rename( Sum_of_Sqs = SumOfSqs,  p_value= `Pr(>F)`) 

#write.csv(perm2324_df, "Table SI Permanova2023_2024.csv", row.names = FALSE)
```



```{r,results='hide', fig.show='hide'}

#build an interaction term to see if there are significant disperions between all the groupings
df_34b$inter_group<-interaction(df_34b$management_trt, df_34b$herbivory_trt, df_34b$year, sep="_")

perm_inter<-betadisper(cov_bray_dis34b, group = df_34b$inter_group)
permutest(perm_inter, permutations = 999)
#this was not significant* 


```



#### Visualization 2023-2024 West pasture

```{r}

#extract nmds scores using site scores
nmds_df34b<-as.data.frame(scores(Tot_NMDS34b, display = "sites"))

#add in the treatments factors from env data frame
nmds_df34b$management_trt<-env_34b$management_trt
nmds_df34b$year<-env_34b$year
nmds_df34b$herbivory_trt<-env_34b$herbivory_trt
nmds_df34b$bison_pasture<-env_34b$bison_pasture


##  colors
herbivory_colors <- c("SB" = "#440154FF", "SX" = "#35B779FF", "XX" = "#31688EFF")


#calculate the centroids and SE for each group then we will multiply by 1.96 to get ~95%CI
centroids_df34b<-nmds_df34b %>% group_by(management_trt, herbivory_trt,year) %>% 
  summarise( NMDS1_mean =mean(NMDS1),
             NMDS2_mean=mean(NMDS2),
             NMDS1_se=sd(NMDS1)/sqrt(n()), 
             NMDS2_se=sd(NMDS2)/sqrt(n()), 
             .groups = "drop")


#now we need trajectory to follow the centroids between years
centroid_trajectory34b<-centroids_df34b %>% arrange(management_trt,herbivory_trt, year) %>% group_by(management_trt,herbivory_trt) %>% mutate(NMDS1_next =lead(NMDS1_mean), 
          NMDS2_next = lead(NMDS2_mean),
          year_next=lead(year), 
          .groups="drop")
         
centroid_trajectory34b<-centroid_trajectory34b %>% 
          filter(!is.na(NMDS1_next)) # keeps only those with next step

centroids_df34b$year<-as.factor(centroids_df34b$year)
centroid_trajectory34b$year<-as.factor(centroid_trajectory34b$year)


  ggplot(centroids_df34b, aes(x=NMDS1_mean, y=NMDS2_mean, color=herbivory_trt, shape = year))+
    facet_wrap(~management_trt)+
          #geom_point(data = nmds_df34b, aes(x=NMDS1, y=NMDS2, color=herbivory_trt), size=2, alpha=0.2)+ #this adds in all plots but looks very messy, keep out for now
           geom_point(size=4, alpha=1)+ #adds my centroids
           geom_errorbar(aes( x= NMDS1_mean, ymin=NMDS2_mean-1.96*NMDS2_se, ymax = NMDS2_mean+1.96*NMDS2_se, color=herbivory_trt), width=0.05, alpha=0.8, linewidth=0.4) + 
           geom_errorbarh(aes(y=NMDS2_mean, xmin=NMDS1_mean -1.96*NMDS1_se, xmax= NMDS1_mean+1.96*NMDS1_se, color=herbivory_trt), height=0.05, alpha=0.8, linewidth=0.4)+
           geom_segment(data = centroid_trajectory34b, aes(x=NMDS1_mean, y=NMDS2_mean, xend=NMDS1_next, yend=NMDS2_next, color=herbivory_trt), arrow = arrow(length=unit(0.15, "cm")), size=1, alpha=0.8)+ #adds our trajectory
           scale_color_manual(values=herbivory_colors,labels= c("SB"= "Bison + Small Mammals", "SX"= "Small Mammals", "XX"= "No Mammals"))+
           scale_shape_manual(values = c(15,16))+
           labs( x= "NMDS 1", y= "NMDS 2", color= "Consumer Treatment", shape= "Year")+ 
           theme_minimal()+
    guides(color=guide_legend(nrow = 1, byrow=TRUE), 
           shape=guide_legend(nrow = 1, byrow=TRUE))+
           theme (legend.position= "bottom", legend.box = "vertical", plot.margin = margin(t=10, r=35, b=10, l=10))+
  coord_cartesian(clip = "off")+
          text_theme(base_size = 12)
  
 ggsave("Ch2_FigureS_2023_2024NMDS.png", width=7.3, height=6.5,units="in", dpi=600)
 
  
```

## 2021 NMDS West

### Data Wrangling
```{r}

# step 1: subset data for only west pasture 
df_W21<- df_wide %>% filter(bison_pasture== "yes") %>% filter(year=="2021")


# and remove columns to create cover df and env df 
cover_W21<- df_W21[, -c(2:9)] # keep unique ID for merging purposes later on

env_W21 <- df_W21[, -c(7:83)] # may need to drop year/block/plot if using envfit (will show up on plots)

```

```{r}
# step 2: standardize the community matrix using site total with decostand

covTotal_W21 <- decostand(cover_W21 [, -c(1)], "total") # dividies by margin total and removes first column 

```

```{r, results='hide', fig.show='hide'}
 # step 3: check dimensions

dimcheckMDS(covTotal_W21, distance = "bray", k=6, trymax = 50, autotransform = FALSE)
#stress test for 2 is great (0.12)
```

```{r, results='hide'}
#step 4: run metaMDS 

Tot_NMDS_W21<- metaMDS(covTotal_W21, distance = "bray", k=2, trymax = 500, maxit=500, autotransform = FALSE, expand = FALSE)

#best solution repeated 2x

```


```{r}
#step 5: standardize the data for dissimilairty distnaces using the data that was standardized by site total baove with decostand 

cov_bray_dis_W21<-vegdist(covTotal_W21, "bray")

# make sure original data frame (subset matrix with env) treatments are all factors 
df_W21$management_trt<-as.factor(df_W21$management_trt)
df_W21$herbivory_trt<- as.factor(df_W21$herbivory_trt)

```

```{r, results='hide'}
#step 6: run permanova with adonis2
permW21<-adonis2(cov_bray_dis_W21~management_trt, permutations = 999, data = df_W21)
permW21
#no significant differences overall p=.291
```



Table S #. PERMANOVA results from plant community composition in the bison  pasture in 2021. 
```{r}
# extract permanova results for table and write as csv

permW21_df<- as.data.frame(permW21) %>% rownames_to_column("Term") %>% 
  rename( Sum_of_Sqs = SumOfSqs,  p_value= `Pr(>F)`) 

#write.csv(permW21_df, "Table SI Permanova2021.csv", row.names = FALSE)
```





